---
title: "Checkpoint 2: Counting Events"
date-modified: "today"
categories: [""]
format:
  html:
    link-external-newwindow: true
    toc: true
  pdf: default
editor_options:
  chunk_output_type: console
---

```{css, echo=FALSE}
.title{
  font-size: 30px;
  color: #006a4e;
  background-color: #fff;
  padding: 0px;
}
```

```{r setup, include=FALSE}
library(here)
source(here("class_dates.R"))
```

::: callout-important
**Due `r cp2 %>% format("%m/%d")`** (12 points). See [Project Central](/project/project_central.qmd) for the full project overview. Make sure you have addressed any feedback from Checkpoint 1.
:::

## Goal

Model a **count** in your scenario with a discrete random variable, and use it to answer a practical question, such as staffing, capacity, or how often something unusual happens.

**Course topics used:** discrete random variables, PMFs and CDFs, expected value and variance, named discrete distributions (Bernoulli, Binomial, Geometric, Poisson, etc.).

## Tasks

### 1. Define your random variable

-   Define $X$ in words, including its units and time frame (e.g., "$X$ = number of patients arriving at the ED between 6 and 7 pm").
-   List the possible values (the support) of $X$.

### 2. Choose and justify a distribution

-   Choose a named distribution for $X$ and state its parameter(s), using a cited value.
-   List the assumptions of that distribution (e.g., fixed number of trials, independence, constant probability or rate), and explain whether each is reasonable for your scenario.
-   Identify the assumption **most likely to fail** in real life, and explain why.

### 3. Solve it analytically

-   Find $E(X)$, $\text{Var}(X)$, and $\text{SD}(X)$ using formulas from class.
-   Calculate **two** probabilities that answer practical questions, for example:
    -   $P(X \geq k)$ for a staffing or capacity threshold $k$, and
    -   $P(X = 0)$ or $P(a \leq X \leq b)$.
-   Show the formula (PMF or CDF) for each, then compute with R (e.g., `dpois()`, `ppois()`, `pbinom()`).

### 4. Simulate it

-   Generate at least 10,000 draws of $X$ with R (e.g., `rpois()`, `rbinom()`, `rgeom()`).
-   Make a plot comparing the **empirical** distribution (from your simulation) with the **theoretical PMF**, for example side-by-side bars or points on a bar chart.
-   Compare the simulated mean, variance, and your two probabilities with the analytic values in a small table.

Starter code (Poisson example; change it for your distribution):

```{r}
#| eval: false
set.seed(550)
lambda <- 12      # mean arrivals per hour (cite your source!)
n_sim  <- 10000

x <- rpois(n_sim, lambda)

tibble::tibble(
  quantity  = c("Mean", "Variance", "P(X >= 18)"),
  analytic  = c(lambda, lambda, 1 - ppois(17, lambda)),
  simulated = c(mean(x), var(x), mean(x >= 18))
)
```

::: callout-warning
## Watch the parameterization

R's `rgeom()` counts the number of **failures before** the first success, which may not match the definition we used in class. Check your definitions, and adjust if needed (e.g., `rgeom(n, p) + 1`).
:::

### 5. Stretch (optional, +1 bonus)

Break the assumption you identified in Task 2 and show how the answer changes. For example, if arrivals are more variable than a Poisson allows, simulate from a model where the rate itself varies from hour to hour, and compare $P(X \geq k)$ with your original answer.

### 6. Say it

Write a short paragraph (4–6 sentences) **for a manager or clinician** who needs to plan around this count. What is typical? What is unusual, and how often does it happen? What would you warn them about, given the assumption most likely to fail?

## Submission

Submit a rendered HTML file (use `code-fold: true`) and your `.qmd`. One submission per team.

## Rubric (12 points)

| Criterion | 3 | 2 | 1 |
|----|----|----|----|
| **Setup** (random variable, distribution, assumptions) | Clear and correct; assumptions discussed with care | Minor gaps | Major errors or missing |
| **Analytic solution** (mean, variance, two probabilities) | Correct, with formulas shown | Small errors | Mostly incorrect or no work shown |
| **Simulation** (draws, PMF plot, comparison table) | Agrees with analytic answers; plot is clear | Runs, but comparison or plot is weak | Missing or incorrect |
| **Interpretation** (paragraph for a manager or clinician) | Accurate, practical, and in plain language | Somewhat technical or vague | Missing or misleading |
