---
title: "Checkpoint 1: When Is a Positive Really Positive?"
date-modified: "today"
categories: [""]
format:
  html:
    link-external-newwindow: true
    toc: true
  pdf: default
editor_options:
  chunk_output_type: console
---

```{css, echo=FALSE}
.title{
  font-size: 30px;
  color: #006a4e;
  background-color: #fff;
  padding: 0px;
}
```

```{r setup, include=FALSE}
library(here)
source(here("class_dates.R"))
```

::: callout-important
**Due `r cp1 %>% format("%m/%d")`** (12 points). See [Project Central](/project/project_central.qmd) for the full project overview. Make sure you have addressed any feedback from Checkpoint 0.
:::

## Goal

Use conditional probability and Bayes' rule to answer: *If someone in my scenario gets a positive result, how likely is it that they truly have the condition?*

**Course topics used:** events, conditional probability, the Law of Total Probability, Bayes' rule, independence.

## Tasks

### 1. Set up the problem

-   Define your events with notation (e.g., $D$ = has the disease, $T^+$ = tests positive).
-   Write each input value from Checkpoint 0 as a probability statement, for example $P(T^+ \mid D) = 0.90$. Cite your sources.
-   Draw a tree diagram **or** a 2×2 table for a hypothetical population of 100,000 people.

### 2. Solve it analytically

-   Use the Law of Total Probability to find $P(T^+)$.
-   Use Bayes' rule to find the **positive predictive value** $P(D \mid T^+)$ and the **negative predictive value** $P(D^c \mid T^-)$.
-   Show your work. You may use R as a calculator, but write out each formula first.

### 3. Simulate it

-   Simulate a population of at least 100,000 people in R. Assign each person a disease status, then a test result that depends on their disease status.
-   Estimate PPV and NPV from your simulated population and compare them with your analytic answers.
-   In 1–2 sentences, explain why the simulated and analytic values are close but not identical.

Starter code (replace the input values with your own):

```{r}
#| eval: false
set.seed(550)
N    <- 1e5
prev <- 0.01   # P(D)
sens <- 0.90   # P(T+ | D)
spec <- 0.95   # P(T- | D^c)

disease  <- rbinom(N, size = 1, prob = prev)
test_pos <- ifelse(disease == 1,
                   rbinom(N, size = 1, prob = sens),
                   rbinom(N, size = 1, prob = 1 - spec))

ppv_sim   <- mean(disease[test_pos == 1])
ppv_bayes <- (sens * prev) / (sens * prev + (1 - spec) * (1 - prev))

c(simulated = ppv_sim, analytic = ppv_bayes)
```

### 4. Explore: how does prevalence matter?

-   Compute the analytic PPV for prevalence values from 0.1% to 50%, keeping sensitivity and specificity fixed.
-   Plot PPV (y-axis) against prevalence (x-axis), and mark your scenario's prevalence on the plot.
-   In 2–3 sentences, describe what happens to PPV as prevalence changes. When would this test be most useful?

### 5. Say it

Write a short paragraph (4–6 sentences) **for a patient** who just received a positive result. Explain what the result means using natural frequencies (e.g., "Out of 1,000 people like you who test positive, about \_\_\_ actually have...") rather than probability notation.

## Submission

Submit a rendered HTML file (use `code-fold: true`) and your `.qmd`. One submission per team.

## Rubric (12 points)

| Criterion | 3 | 2 | 1 |
|----|----|----|----|
| **Setup** (notation, input values, tree or table) | Clear and correct; inputs cited | Minor gaps in notation or citation | Major errors or missing |
| **Analytic solution** (LTP, PPV, NPV) | Correct, with work shown | Small errors | Mostly incorrect or no work shown |
| **Simulation** (code, comparison to analytic answer, prevalence plot) | Agrees with analytic answer; differences explained; plot is clear | Runs, but comparison or plot is weak | Missing or incorrect |
| **Interpretation** (prevalence discussion, patient paragraph) | Accurate and in plain language | Somewhat technical or vague | Missing or misleading |
