---
title: "Checkpoint 3: Averages and the Central Limit Theorem"
date-modified: "today"
categories: [""]
format:
  html:
    link-external-newwindow: true
    toc: true
  pdf: default
editor_options:
  chunk_output_type: console
---

```{css, echo=FALSE}
.title{
  font-size: 30px;
  color: #006a4e;
  background-color: #fff;
  padding: 0px;
}
```

```{r setup, include=FALSE}
library(here)
source(here("class_dates.R"))
```

::: callout-important
**Due `r cp3 %>% format("%m/%d")`** (12 points). See [Project Central](/project/project_central.qmd) for the full project overview. Make sure you have addressed any feedback from Checkpoint 2.
:::

## Goal

Model a **time or measurement** in your scenario with a continuous random variable, then use the Central Limit Theorem to answer a question about an **average** of many observations.

**Course topics used:** continuous random variables, PDFs and CDFs, quantiles, named continuous distributions (Uniform, Exponential, Gamma, Normal), sums of independent random variables, the Central Limit Theorem.

## Tasks

### 1. Define your random variable

-   Define $Y$ in words, including its units (e.g., "$Y$ = time in hours from ED arrival to being seen by a provider").
-   State the support of $Y$.

### 2. Choose and justify a distribution

-   Choose a named continuous distribution for $Y$, and state its parameter(s) using cited values.
-   Explain why this distribution is a reasonable choice. Consider its shape (symmetric or skewed?), its support, and what we know about the real quantity.
-   Plot the PDF of your chosen distribution.

### 3. Solve it analytically

-   Find $E(Y)$ and $\text{Var}(Y)$.
-   Calculate **two** probabilities or quantiles that answer practical questions, for example:
    -   $P(Y > c)$ for a clinically meaningful cutoff $c$, and
    -   the 90th percentile of $Y$.
-   Show the setup (integral or CDF) for at least one of them, then compute with R (e.g., `pexp()`, `qgamma()`, `pnorm()`).

### 4. Simulate the Central Limit Theorem

Let $\bar{Y}_n$ be the average of $n$ independent observations of $Y$.

-   For $n = 5$, $30$, and $100$, simulate at least 5,000 sample means.
-   For each $n$, plot a histogram of the sample means with the normal curve predicted by the CLT, $N\left(\mu, \sigma^2/n\right)$, overlaid.
-   Compare the mean and SD of your simulated sample means with the CLT's predictions.
-   In 2–3 sentences, describe how quickly the distribution of $\bar{Y}_n$ looks normal. How does the shape (skewness) of $Y$ affect this?

Starter code (Exponential example; change it for your distribution):

```{r}
#| eval: false
set.seed(550)
rate  <- 1 / 2.5          # mean wait of 2.5 hours (cite your source!)
mu    <- 1 / rate
sigma <- 1 / rate
n_rep <- 5000

sim_means <- function(n) replicate(n_rep, mean(rexp(n, rate)))

means_30 <- sim_means(30)

hist(means_30, freq = FALSE, breaks = 40,
     main = "Sample means, n = 30", xlab = "Average wait (hours)")
curve(dnorm(x, mean = mu, sd = sigma / sqrt(30)), add = TRUE, lwd = 2)
```

### 5. Answer a question about an average

Use the CLT to answer a practical question about $\bar{Y}_n$ with an $n$ that makes sense for your scenario. For example: *What is the probability that the average wait of the next 50 patients is more than 3 hours?*

-   Compute the CLT approximation.
-   Check it with your simulation.
-   Compare it with the probability that a **single** observation exceeds the same cutoff, and explain why the two answers differ.

### 6. Say it

Write a short paragraph (4–6 sentences) **for a clinician or administrator** that explains the difference between what happens to *one* patient and what happens *on average* across many patients, using your results.

## Submission

Submit a rendered HTML file (use `code-fold: true`) and your `.qmd`. One submission per team.

## Rubric (12 points)

| Criterion | 3 | 2 | 1 |
|----|----|----|----|
| **Setup** (random variable, distribution, justification, PDF plot) | Clear and correct; choice well justified | Minor gaps | Major errors or missing |
| **Analytic solution** (mean, variance, probabilities or quantiles, CLT approximation) | Correct, with work shown | Small errors | Mostly incorrect or no work shown |
| **Simulation** (CLT histograms for three sample sizes, comparison with predictions) | Clear plots; matches CLT predictions; role of skewness explained | Runs, but comparison is weak | Missing or incorrect |
| **Interpretation** (one patient vs. the average) | Accurate and in plain language | Somewhat technical or vague | Missing or misleading |
