---
title: "Checkpoint 1: When Is a Positive Really Positive?"
author: "Your name(s) here!"
subtitle: "BSTA 550"
date-modified: "today"
format:
  html:
    link-external-newwindow: true
    toc: true
    embed-resources: true
    number-sections: true
    number-depth: 3
    code-fold: true
editor_options:
  chunk_output_type: console
---

```{r}
#| message: false
#| echo: false

# PLEASE DO NOT REMOVE THIS CODE CHUNK!!!
### ADD YOUR LIBRARIES HERE!!! ####
library(tidyverse)
```

::: callout-warning
Note that this is the editing file. **There are detailed instructions on our course page for Checkpoint 1 that are not in this file.**

-   Please do not remove anything from this file. Only add your work to it.
-   Name your files `lastname_firstinitial_checkpoint_01.qmd` and `.html` (or `lastname1_lastname2_checkpoint_01.qmd` and `.html` for a pair). One submission per team.
:::

## Your scenario {.unnumbered}

Scenario (A, B, or C):

## Set up the problem

::: callout-important
#### Task

-   Copy your events and your three Checkpoint 1 probability statements from Checkpoint 0, with any corrections from feedback. Keep the source for each value.
-   Add the one conditional probability you did not record in Checkpoint 0, using the complement rule: $P(T^+ \mid D^c)$ for Scenarios A and B, or $P(T^- \mid D^c)$ for Scenario C.
-   Draw a tree diagram **or** a 2×2 table for a hypothetical population of 100,000 people screened by your program. Round each count to a whole number of people.
:::

**Events:**

-   $D$ =
-   $T^+$ =

| Input | Probability statement | Source |
|----|----|----|
| Prevalence |  |  |
| Sensitivity |  |  |
| Specificity (A, B) or false-positive rate (C), from Checkpoint 0 |  |  |
| New: complement of the row above |  | Complement rule |

**Tree diagram or 2×2 table:**

## Solve it analytically

::: callout-important
#### Task

-   Use the Law of Total Probability to find $P(T^+)$.
-   Use Bayes' rule to find the positive predictive value $P(D \mid T^+)$ and the negative predictive value $P(D^c \mid T^-)$.
-   Write out each formula first, then compute (you may use R as a calculator).
:::

```{r}

```

## Simulate it

::: callout-important
#### Task

-   Simulate a population of at least 100,000 people. Assign each person a disease status, then a test result that depends on their disease status.
-   Estimate PPV and NPV from your simulation and compare them with your analytic answers.
-   In 1–2 sentences, explain why the simulated and analytic values are close but not identical.
:::

```{r}

```

## Explore: how does prevalence matter?

::: callout-important
#### Task

-   Compute the analytic PPV for prevalence values from 0.01% to 50%, keeping sensitivity and specificity fixed.
-   Plot PPV (y-axis) against prevalence (x-axis), and mark your scenario's prevalence.
-   In 2–3 sentences, describe what happens to PPV as prevalence changes. Then describe a group or setting your program might serve with a higher or lower prevalence, and how a positive result would mean something different there.
:::

```{r}

```

## Say it

::: callout-important
#### Task

-   Write a short paragraph (4–6 sentences) for the letter or text message your program sends to every person with a positive result. Use natural frequencies instead of probability notation, and say what the next step is.
-   Answer your Checkpoint 1 guiding question from Checkpoint 0 in one sentence.
:::

**Message:**

**Answer to my Checkpoint 1 guiding question:**
