---
title: "Checkpoint 3: Averages and the Central Limit Theorem"
author: "Your name here!"
subtitle: "BSTA 550"
date-modified: "today"
format:
  html:
    link-external-newwindow: true
    toc: true
    embed-resources: true
    number-sections: true
    number-depth: 3
    code-fold: true
editor_options:
  chunk_output_type: console
---

```{r}
#| message: false
#| echo: false

# PLEASE DO NOT REMOVE THIS CODE CHUNK!!!
### ADD YOUR LIBRARIES HERE!!! ####
library(tidyverse)
```

::: callout-warning
Note that this is the editing file. **There are detailed instructions on our course page for Checkpoint 3 that are not in this file.**
:::

## Your scenario {.unnumbered}

Scenario (A, B, or C):

## Define your random variable

::: callout-important
#### Task

-   Write $Y$ in words, including its units, using the definition for your scenario.
-   State the support of $Y$.
:::

## Justify the distribution

::: callout-important
#### Task

-   State the distribution for your scenario and its parameter(s), citing where each value came from.
-   Explain why this distribution is a reasonable choice, and where it might not match the real quantity.
-   Plot the PDF of your distribution.
:::

```{r}

```

## Solve it analytically

::: callout-important
#### Task

-   Find $E(Y)$ and $\text{Var}(Y)$.
-   Calculate **two** probabilities or quantiles that answer practical questions for your program. Show the setup (integral or CDF) for at least one of them, then compute with R.
:::

```{r}

```

## Simulate the Central Limit Theorem

::: callout-important
#### Task

-   For $n = 5$, $30$, and $100$, simulate at least 5,000 sample means.
-   For each $n$, plot a histogram of the sample means with the normal curve predicted by the CLT overlaid.
-   Compare the mean and SD of your simulated sample means with the CLT's predictions.
-   In 2–3 sentences, describe how quickly the distribution of $\bar{Y}_n$ looks normal, and how the shape (skewness) of $Y$ affects this.
:::

```{r}

```

## Answer a question about an average

::: callout-important
#### Task

-   Use the CLT to answer a practical question about $\bar{Y}_n$, using the $n$ for your scenario.
-   Check it with your simulation.
-   Compare it with the probability that a **single** observation is beyond the same cutoff, and explain why the two answers differ.
:::

```{r}

```

## Say it

::: callout-important
#### Task

Write a short paragraph (4–6 sentences) for the program manager or health department that explains the difference between what happens to *one* person and what happens *on average* across many people. Why does this difference matter when the program reports averages to the public?
:::
