---
title: "Checkpoint 1: When Is a Positive Really Positive?"
author: "Your name here!"
subtitle: "BSTA 550"
date-modified: "today"
format:
  html:
    link-external-newwindow: true
    toc: true
    embed-resources: true
    number-sections: true
    number-depth: 3
    code-fold: true
editor_options:
  chunk_output_type: console
---

```{r}
#| message: false
#| echo: false

# PLEASE DO NOT REMOVE THIS CODE CHUNK!!!
### ADD YOUR LIBRARIES HERE!!! ####
library(tidyverse)
```

::: callout-warning
Note that this is the editing file. **There are detailed instructions on our course page for Checkpoint 1 that are not in this file.**
:::

## Your scenario {.unnumbered}

Scenario (A, B, or C):

## Set up the problem

::: callout-important
#### Task

-   Define your events with notation, using the events for your scenario.
-   Write each input value as a probability statement, and cite the source for each value.
-   Draw a tree diagram **or** a 2×2 table for a hypothetical population of 100,000 people screened by your program.
:::

## Solve it analytically

::: callout-important
#### Task

-   Use the Law of Total Probability to find $P(T^+)$.
-   Use Bayes' rule to find the positive predictive value $P(D \mid T^+)$ and the negative predictive value $P(D^c \mid T^-)$.
-   Write out each formula first, then compute (you may use R as a calculator).
:::

```{r}

```

## Simulate it

::: callout-important
#### Task

-   Simulate a population of at least 100,000 people. Assign each person a disease status, then a test result that depends on their disease status.
-   Estimate PPV and NPV from your simulation and compare them with your analytic answers.
-   In 1–2 sentences, explain why the simulated and analytic values are close but not identical.
:::

```{r}

```

## Explore: how does prevalence matter?

::: callout-important
#### Task

-   Compute the analytic PPV for prevalence values from 0.1% to 50%, keeping sensitivity and specificity fixed.
-   Plot PPV (y-axis) against prevalence (x-axis), and mark your scenario's prevalence.
-   In 2–3 sentences, describe what happens to PPV as prevalence changes. Then describe a group or setting your program might serve with a higher or lower prevalence, and how a positive result would mean something different there.
:::

```{r}

```

## Say it

::: callout-important
#### Task

Write a short paragraph (4–6 sentences) for the letter or text message your program sends to every person with a positive result. Use natural frequencies instead of probability notation, and say what the next step is.
:::
